Effects of bariatric surger and weight loss on resting-state functional connectivity

### 2021-01-07 manuscript/draft_rsfmri.Rmd

library(sjPlot) #only works for HTML output
library(texreg) #works for Rmarkdown
library(knitr) # version 1.30
```

```{r "knit setup", include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE) # fig.height=2, fig.width=4)
options(knitr.table.format = "latex")
options(knitr.kable.NA = '') # hide NA in tables
# set working directory of knit (RMarkdown) to directory of Rproj
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
setwd(knitr::opts_knit$get("root.dir"))
```

@@

```{r "Load data", include=FALSE}
source("report/create_sample_df.R")
final = create_sample_df(group = "all", tp = "all")

@@

```{r "tableSample", echo=FALSE}
source("./report/tables.R")
tab_sample <- mk_SampleTable(final)

@@

<!-- STUDY DESIGN --> The ADIPOSITAS-Study conducted in the Charité Berlin is a observational, prospective and longitudinal design described in detail by @Prehn_2020. Data collection is still ongoing but only data available in April 2019 is used.
The procedures followed were in accordance with the Helsinki Declaration. This study was registered at clinicaltrials.gov as NCT01554228 and described in more detailed in the preregistration under [https://osf.io/mzyxv](https://osf.io/mzyxv).
<!-- RECRUITEMENT, INCLUSION/EXCLUSION CRITERIA --> Participants were recruited from the Center for Bariatric and Metabolic Surgery if they were between age $18$ and $70$ and fulfilled the inclusion criteria. These were, in accordance with BARS guidelines, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013].
Participants with a history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (MMSE score $< 34$) were excluded from the study.
Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery or after the baseline appointment but not all participants have complete data for all time points (see Table \@ref(tab:tableDescr)). `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for only one time point.
The current sample is the subsample of $69$ subjects that were originally enrolled in the study entailing only subjects that have received MRI. Analyses were performed on all participants that provided at least one data point of fMRI data (detailed information on data availability in \@ref(tab:tableSample)).

<!-- DESCRIPTIVE STATISTICS --> The final sample entailed morbidly obese individuals (n = `r nrow(subj_df)`, `r nrow(filter(subj_df,Sex == "f"))` female; ages `r round(mean(subj_df$Age_BL),2)` $\pm$ `r round(sd(subj_df$Age_BL),2)` SD years (range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`), that either underwent surgery (n = `r cs["Sum","KG"]`, `r cs["f","IG"]` female) or were waiting list controls (n = `r cs["Sum","KG"]`, `r cs["f","KG"]` female).
The study design was quasi-experimental, as patient in the waiting control list had also been recommended to undergo bariatric surgery, but were still waiting for their health insurance's approval.
```{r tableBaselineCharacteristics, include=FALSE}
dfBL <- create_sample_df(group = "all", tp = "BL")

@@

```
<!-- STUDY PROCEDURE --> Subjects arrived in the morning (between 07:00 and 12:00) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropometric measurements before having a one our break for <!-- standadized??  -->breakfast. MRI scanning was done after further psychological tests (for details see @Prehn_2020).
```{r figBMIdescr, fig.align='center', out.width="90%", fig.cap="BMI over time.", echo=FALSE}
# eval = TRUE, fig.asp = 0.7, fig.width = 6
source("report/figures.R")
fig_BMIdescr <- mk_figBMIdescr(final)
fig_BMIdescr
```

@@

### Global Signal regression   
[@Ciric_2017] showed that GSR is very efficient in removing the correlation of head motion and FC, and outperforms ICA-AROMA alone. Yet, it introduces spatial dependency and spurious correlations. Previously, we reported on head motion differences between groups over time in this study [@Beyer_2020], and thus it was very important to minimize the motion-FC correlation. The global signal was derived from the average of all voxels in the brain mask. Then, we regressed WM and CSF CompCor components along with the GS from the ICA-AROMA fMRI data and high-pass filtered the resulting file as above.
*In the preregistration we stated " As global signal regression is very efficient in removing the correlation of head motion and connectivity dependence, but introduces spatial dependency and spurious correlations, we perform GSR after ICA-AROMA in a separate sensitivity analysis."*

@@

<!-- DESIGN MATRIX, vielleicht beide auch als Abbildung kurz zeigen? --> For the first model, time was introduced as factor because we did not assume strict linearity for the increase in connectivity and therefore refrained from modeling time as continuous variable (see Figure \@ref(fig:figDesignmatrix)). The resulting factorial design <!-- check: flexible factorial? --> contained one regressor for each time point per group.
```{r figDesignmatrix, fig.align = 'center', fig.ncol = 2, out.width = "50%", fig.subcap=c('Design matrix with time as categorial factor.','Design matrix with time as continuous variable.'), fig.cap="Possible design matrices of time-by-group model.", echo=FALSE}
# fig.show="hold",  out.width="40%", fig.height = 8,
source("report/figures.R")
PlotList <- mk_figDesignMatrix()

@@

Figure \@ref{fig:figNetworks} shows the DMN and reward network which we used for extracting the aggregated FC values. Both networks included the typically described brain regions (i.e. medial prefrontal cortex, parietal lobules and posterior hippocampus for the DMN, and anterior cingulare, right amygdala and ventromedial-prefrontal cortex for the reward network).

```{r prepare tsnr ROI plot, echo=FALSE}
final$subj.ID_tp=paste0(final$subj.ID,'_', final$tp)

tsnr=read.table("/data/pt_02161/Analysis/Preprocessing/qa/rs_qa/group_level_QA/check_tsnr/tsnr_in_ROIs.txt")
tsnr=tsnr[tsnr$V1!="subj",]
colnames(tsnr)=c("subj.ID_tp","roi","median","mean","sd")
levels(tsnr$roi)=droplevels(tsnr$roi)
levels(tsnr$roi)=c("NAcc", "precuneus")
tsnr$mean=as.double(tsnr$mean)

final_tsnr=merge(final, tsnr, by="subj.ID_tp", all.x=TRUE)

p<-ggplot(aes(as.factor(roi), mean),data=final_tsnr) +
  geom_violin(aes(fill=roi)) + geom_jitter(height = 0, width = 0.1) + xlab("Region of interest") + ylab("average tSNR") +
  scale_fill_manual(values=wes_palette("Darjeeling2",4,type="discrete")[2:3]) +
  theme(axis.text=element_text(size=10),
  axis.title=element_text(size=12),
  strip.text = element_text(size=12),
  legend.position="")
```

```{r figtSNR,fig.cap="Lower tSNR in the region of interest NAcc versus precuneus used for seed-based connectivity (shown are all 101 timepoints)", out.width="60%", fig.align="center", echo=FALSE}
p
```
Yet, the reward network was less pronounced (overall lower T-values) and symmetric (no significant connectivity with left amygdala), which might be due to lower SNR in the seeded brain region (see Figure \@ref{fig:figtSNR}.

@@

```{r tableBMImodel}
source("report/tables.R")
TableList <- mk_FCTables()
tab_BMImodel <- TableList[[1]]
tab_BMImodel
```

>>>>

library(sjPlot) #only works for HTML output
library(texreg) #works for Rmarkdown
library(knitr) # version 1.30

```{r "knit setup", include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE) # fig.height=2, fig.width=4)
options(knitr.table.format = "latex")
options(knitr.kable.NA = '') # hide NA in tables
# set working directory of knit (RMarkdown) to directory of Rproj
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
setwd(knitr::opts_knit$get("root.dir"))

source("report/figures.R")
source("report/tables.R")
source("report/create_sample_df.R")
```

@@

```{r "Load data", include=FALSE}
final = create_sample_df(group = "all", tp = "all")

@@

```{r "tableSample", echo=FALSE}
tab_sample <- mk_SampleTable(final)

@@

<!-- STUDY DESIGN --> The ADIPOSITAS-Study conducted in the Charité Berlin is a observational, prospective and longitudinal design described in detail by @Prehn_2020. Data collection is still ongoing but only data available in April 2019 is used.
The procedures followed were in accordance with the Helsinki Declaration. This study was registered at clinicaltrials.gov as NCT01554228 and described in more detailed in the preregistration under [https://osf.io/yp42s](https://osf.io/yp42s). <!-- previous version: https://osf.io/f8tpn, update: https://osf.io/59bh7/ -->
<!-- RECRUITEMENT, INCLUSION/EXCLUSION CRITERIA --> Participants were recruited from the Center for Bariatric and Metabolic Surgery if they were between age $18$ and $70$ and fulfilled the inclusion criteria. These were, in accordance with BARS guidelines, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013].
Participants with a history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (MMSE score $< 34$) were excluded from the study.
Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery or after the baseline appointment but not all participants have complete data for all time points (see Table \@ref(tab:tableDescr)). `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for only one time point.
The current sample is the subsample of $69$ subjects that were originally enrolled in the study entailing only subjects that have received MRI. Analyses were performed on all participants that provided at least one data point of fMRI data (detailed information on data availability in \@ref(tab:tableSample)).

<!-- DESCRIPTIVE STATISTICS --> The final sample entailed morbidly obese individuals 
(n = `r nrow(subj_df)`, 
`r nrow(filter(subj_df,Sex == "f"))` female; 
`r round(mean(subj_df$Age_BL),2)` 
$\pm$ `r round(sd(subj_df$Age_BL),2)` SD years old
(range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`), 
that either underwent surgery 
(n = `r cs["Sum","IG"]`, 
`r cs["f","IG"]` female, 
`r round(mean(subj_df$Age_BL[subj_df$condition == "IG"]),2)` 
$\pm$ `r round(sd(subj_df$Age_BL[subj_df$condition == "IG"]),2)` SD years old) 
or were waiting list controls 
(n = `r cs["Sum","KG"]`, 
`r cs["f","KG"]` female,
`r round(mean(subj_df$Age_BL[subj_df$condition == "KG"]),2)`
$\pm$ `r round(sd(subj_df$Age_BL[subj_df$condition == "KG"]),2)` SD years old) 
The study design was quasi-experimental, as patient in the waiting control list had also been recommended to undergo bariatric surgery, but were still waiting for their health insurance's approval.
```{r tableBaselineCharacteristics, include=FALSE}
dfBL <- create_sample_df(group = "all", tp = "BL")

@@

```
<!-- STUDY PROCEDURE --> Subjects arrived in the morning (between 07:00 and 12:00) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropometric measurements before having a one our break for <!-- standadized??  -->breakfast. MRI scanning was done after further psychological tests (for details see @Prehn_2020).
```{r figBMIdescr, fig.align='center', out.width="90%", fig.cap="BMI over time.", echo=FALSE}
# eval = TRUE, fig.asp = 0.7, fig.width = 6
fig_BMIdescr <- mk_figBMIdescr(final)
fig_BMIdescr
```

@@

### Global Signal regression   
@Ciric_2017 showed that GSR is very efficient in removing the correlation of head motion and FC, and outperforms ICA-AROMA alone. Yet, it introduces spatial dependency and spurious correlations. Previously, we reported on head motion differences between groups over time in this study [@Beyer_2020], and thus it was very important to minimize the motion-FC correlation. The global signal was derived from the average of all voxels in the brain mask. Then, we regressed WM and CSF CompCor components along with the GS from the ICA-AROMA fMRI data and high-pass filtered the resulting file as above.
*In the preregistration we stated " As global signal regression is very efficient in removing the correlation of head motion and connectivity dependence, but introduces spatial dependency and spurious correlations, we perform GSR after ICA-AROMA in a separate sensitivity analysis."*

@@

<!-- DESIGN MATRIX, vielleicht beide auch als Abbildung kurz zeigen? --> For the first model, time was introduced as factor because we did not assume strict linearity for the increase in connectivity and therefore refrained from modeling time as continuous variable (see Figure \@ref(fig:figDesignmatrix)). The resulting factorial design <!-- check: flexible factorial? --> contained one regressor for each time point per group.
```{r figDesignmatrix, fig.align = 'center', fig.ncol = 2, out.width = "50%", fig.subcap=c('Design matrix with time as categorial factor.','Design matrix with time as continuous variable.'), fig.cap="Possible design matrices of time-by-group model.", echo=FALSE}
# fig.show="hold",  out.width="40%", fig.height = 8,
PlotList <- mk_figDesignMatrix()

@@

Figure \@ref{fig:figNetworks} shows the DMN and reward network which we used for extracting the aggregated FC values. Both networks included the typically described brain regions (i.e. medial prefrontal cortex, parietal lobules and posterior hippocampus for the DMN, and anterior cingulate, right amygdala and ventromedial-prefrontal cortex for the reward network).

```{r prepare tsnr ROI plot, echo=FALSE}
fig_tSNR <- mk_figtSNR(final)
```

```{r figtSNR,fig.cap="Lower tSNR in the region of interest NAcc versus precuneus used for seed-based connectivity (shown are all 101 timepoints)", out.width="60%", fig.align="center", echo=FALSE}

fig_tSNR
```
Yet, the reward network was less pronounced (overall lower T-values) and symmetric (no significant connectivity with left amygdala), which might be due to lower SNR in the seeded brain region (see Figure \@ref{fig:figtSNR}.

@@

```{r tableBMImodel}
TableList <- mk_FCTables()
tab_BMImodel <- TableList[[1]]
tab_BMImodel
```

### 2021-01-07 manuscript/intro_rsfmri.Rmd

---
title: "Effects of bariatric surger and weight loss on resting-state functional connectivity"
author:
- xxx [1]
- xxx [1][2]
- xxx [1][2][3]
- xxx [1][2]
date: 1 Max-Planck-Institute for Human Cognitive and Brain Sciences, Leipzig \newline
  2 CRC 1052 "Obesity Mechanisms", Subproject A1, University of Leipzig \newline 3
  Day Clinic for Cognitive Neurology, University Clinic Leipzig
output: word_document
editor_options:
  chunk_output_type: console
bibliography: bibliography.bib
---
```{r "load packages", include=FALSE}

@@

```{r "knit setup"}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE) # fig.height=2, fig.width=4)
options(knitr.table.format = "latex") 
# set working directory of knit (RMarkdown) to directory of Rproj
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
setwd(knitr::opts_knit$get("root.dir"))
```

@@

<!-- novelty statement --> As described previously, several studies have investigated the effects of bariatric surgery on the intrinsic functional connectivity in patients. Yet, to our knowledge there is no quasi-experimental study, comparing a cohort of patients to a cohort with similar characteristics but unexposed to the intervention, i.e. bariatric surgery, over an extended time period as would be demanded to draw more substantiated inference [@George_2016].
<!-- hypotheses --> In this study, we aim to bridge this gap and investigate possible dynamics of within-reward network connectivity before and after bariatric surgery by comparison with a contemporaneously measured waiting-control group. Moreover, we will explore whether the changes in functional connectivity are associated with the amount of weight lost after bariatric surgery treatment in obesity to explore a potential gradient link. In addition, we will investigate potential changes in within-network functional connectivity after surgery in other resting-state networks such as default mode network.
Participants are measured at baseline, at 6 months and at 24 months to capture both phases of rapid weight-loss and maintenance as expected from non-surgical interventions [@Shai_2008]. Notably, we will consider that different courses in increase and decrease of functional connectivity that are not necessarily linear are possible over the course of one year, depending on the phase of weight management [@Olivo_2017].

>>>>

---
title: "Effects of bariatric surger and weight loss on resting-state functional connectivity"
author:
- xxx [1]
- xxx [1][2]
- xxx [1][2][3]
- xxx [1][2]
date: 1 Max-Planck-Institute for Human Cognitive and Brain Sciences, Leipzig \newline
  2 CRC 1052 "Obesity Mechanisms", Subproject A1, University of Leipzig \newline 3
  Day Clinic for Cognitive Neurology, University Clinic Leipzig
output:
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: console
bibliography: bibliography.bib
---
```{r "load packages", include=FALSE}

@@

```{r "knit setup", echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE) # fig.height=2, fig.width=4)
options(knitr.table.format = "latex") 
# set working directory of knit (RMarkdown) to directory of Rproj
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
setwd(knitr::opts_knit$get("root.dir"))
```

@@

<!-- novelty statement --> As described previously, several studies have investigated the effects of bariatric surgery on the intrinsic functional connectivity in patients. Yet, to our knowledge there is no quasi-experimental study, comparing a cohort of patients to a cohort with similar characteristics but unexposed to the intervention, i.e. bariatric surgery, over an extended time period as would be demanded to draw more substantiated inference [@George_2016].
<!-- hypotheses --> In this study, we aim to bridge this gap and investigate possible dynamics of within-reward network connectivity before and after bariatric surgery by comparison with a contemporaneously measured waiting-control group. Moreover, we will explore whether the changes in functional connectivity are associated with the amount of weight lost after bariatric surgery treatment in obesity to explore a potential gradient link. In addition, we will investigate potential changes in within-network functional connectivity after surgery in other resting-state networks such as default mode network.
Participants are measured at baseline, at 6 months and at 12 months to capture both phases of rapid weight-loss and maintenance as expected from non-surgical interventions [@Shai_2008]. Notably, we will consider that different courses in increase and decrease of functional connectivity that are not necessarily linear are possible over the course of one year, depending on the phase of weight management [@Olivo_2017].
